Viscous Erosion in Porous Media

### 2018-01-09 docs/methods/erosion_methods.tex
Worked on writing in theta-L numerical methods. Soon, I plan to write in the results section to explain the area scaling laws, the terminal angle, and the tables.

In order to regularize boundary evolution, we modify the normal velocity
in Eq.~\eqref{eqn:bodyEvolution} as follows
\begin{equation}
\label{eqn:Vn}
\Vn = \atau + \eps \mean{\atau} \left( \frac{L }{2\pi} \kappa - 1 \right)
\end{equation}
Here, $\kappa$ is the curvature, $\eps$ is a small parameter, and $L$ is the total arc length of the body, and $\mean{\cdot}$ indicates a surface mean. The second term on the right acts to smooth bodies with high-frequency oscillations by introducing a {\em curvature}-dependent component in the interface velocity. Since a closed body has a mean curvature of $\mean{\kappa} = 2\pi / L$, the term in parentheses has mean zero. Thus, this term is area-preserving, so that the only material loss is due to the shear stress itself. In addition, we have scaled this term on the $\mean{\atau}$, the average absolute stress, so that it is always proportional to the true source of erosion $\atau$, even as bodies vanish (reword).

We now introduce the tangent angle, $\theta$, defined through the relation
\begin{equation}
\left( \pderiv{x}{s}, \pderiv{y}{s} \right) = \left(\cos \theta, \sin \theta \right)
\end{equation}
where $s$ is arc length. We also introduce the normalized arc length $\alpha = 2 \pi s / L$. In terms of these new variables, the curvature can be expressed as
\begin{equation}
\kappa = \frac{2 \pi}{L} \pderiv{\theta}{\alpha}
\end{equation}
Substitution into Eq.~\eqref{eqn:Vn} gives
\begin{equation}
\label{eqn:Vn2}
\Vn = \atau +  \eps \mean{\atau}   \left(\pderiv{\theta}{\alpha} - 1 \right)
%\Vn = \atau + \frac{2 \pi \eps}{L \log \left(2\pi/L \right)}  \left(\pderiv{\theta}{\alpha} - 1 \right)
\end{equation}

The main idea is to introduce an artificial tangential velocity that keeps collocation points equally spaced with respect to arc length; i.e.~the grid of $\alpha$ should always remain equally spaced. The tangential velocity that does the trick is determined by \cite{hou-low-she1994}
\begin{equation}
\pderiv{\Vs}{\alpha} = \thalpha \Vn - \mean{\thalpha \Vn}
\end{equation}
The interface evolution is then given by
\begin{equation}
\dot{\xx}(t) = \Vn \nn + \Vs \ss.
\end{equation}
Upon transforming to the variables $\theta(\alpha,t)$ and $L(t)$, we obtain the evolution equations
\begin{align}
\label{eqn:Levo0}
& \tderiv{L}{t} = - 2 \pi \mean{\thalpha \Vn} \\
\label{eqn:thetaevo0}
& \pderiv{\theta}{t} = \frac{2 \pi}{L} \left( \pderiv{\Vn}{\alpha} + \thalpha \Vs \right)
\end{align}
Inserting Eq.~\eqref{eqn:Vn2} for the normal velocity gives
\begin{equation}
\pderiv{\theta}{t} = \frac{2\pi}{L} \left(
\eps \mean{\atau} \ppd{\theta}{\alpha} + \pderiv{\atau}{\alpha} + \thalpha \Vs \right)
\end{equation}

For simplicity, we introduce the variables
\begin{align}
\label{eq:M}
& \Mterm = - 2 \pi \mean{\thalpha \Vn} \\
\label{eq:N}
& \NLterm = \frac{2 \pi}{L} \left( \pderiv{\atau}{\alpha} + \thalpha \Vs \right) \\
\label{eq:elfun}
& \elfun = \frac{2 \pi  \mean{\atau}}{L}
%\frac{4\pi^2}{L^{2} \log \left(2\pi/L \right)}
\end{align} 
Then Eq.~\eqref{eqn:thetaevo0} can be rewritten as
\begin{equation}
\label{eqn:thetaevo1}
\pderiv{\theta}{t} = \eps \elfun \ppd{\theta}{\alpha} + \NLterm
\end{equation}
Note that the highest-order, i.e.~stiff, term in Eq.~\eqref{eqn:thetaevo1} is {\em quasi-linear}. This is the key to the {\thL} method, and allows us to use very stable, implicit methods for this stiff term.

@@

%% TIME-STEPPING %%
\subsection{Time-stepping with the {\thL} method} 
We will evolve $\theta$ in spectral space, and so introduce the transforms
\begin{align}
& \theta(\alpha,t) = \alpha + \FourierSum \thhat_k e^{ i k \alpha} \\
& \NLterm(\alpha,t)  = \FourierSum \hat{\NLterm}_k e^{ i k \alpha}
\end{align}
Recall that $\alpha = 2 \pi s / L$ is a normalized arc length.
Equation~\eqref{eqn:thetaevo1} gives a set of ODEs for the coefficients
\begin{equation}
\tderiv{\thhat_k}{t} +  \eps k^2  \elfun \thhat_k = \hat{\NLterm}_k
\end{equation}
We introduce the integrating factor
\begin{equation}
\label{eq:mu}
\mu_k = \exp \left( \eps k^2 \int \elfun \, dt \right)
\end{equation}
Then the evolution equations for $L$ and $\theta$ can be recast as the coupled system of ODEs
\begin{align}
\label{eqn:Levo2}
& \tderiv{L}{t} = \Mterm \\
\label{eqn:thetaevo2}
& \tderiv{}{t}\left( \mu_k \thhat_k \right) = \mu_k \hat{\NLterm}_k
\end{align}
We solve system \eqref{eqn:Levo2}--\eqref{eqn:thetaevo2} with an explicit, second-order Runge-Kutta method (RK2). In particular, we choose the {\em midpoint} RK2 method so as to maximize the smoothing effect of the curvature term.

@@

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[htbp]
\begin{center}
\includegraphics[width = 0.8 \textwidth]{./figs/arangle.pdf}
\caption{Aspect ration and opening angle}
\label{}
\end{center}
\end{figure}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
% Figure from datafile 01circ1024c5
% Uses 01circ1024.in with epsfac = 5, sigfac = 5, dt = 2e-4, fixarea = 1, fixpdrop = 0

% Table
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{table}%[htbp]
\begin{center}
\caption{Final aspect ratio and opening angle
} 
\vspace{0.3 pc}
%\label{aaa}
\begin{tabular}{c c c}
\hline
\hspace{0.5pc} smoothing parameters: $\eps$ and $\sigma$
\hspace{0.5pc} & final aspect ratio 
\hspace{0.5pc} & final opening angle \\
\hline
20/1024		& 2.55	& $110^\circ \pm 6^\circ$	\\
10/1024		& 2.62	& $104^\circ \pm 4^\circ$	\\
5/1024		& 2.65	& $102^\circ \pm 2^\circ$	\\
\hline
\end{tabular}
\end{center}
\end{table}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
% npts = 1024

>>>>

In order to regularize boundary evolution, we modify the normal velocity
in \eqref{eqn:bodyEvolution} as follows
\begin{equation}
\label{eqn:Vn}
\Vn = \atau + \eps \mean{\atau} \left( \frac{L }{2\pi} \kappa - 1 \right)
\end{equation}
Here, $\kappa$ is the curvature, $\eps$ is a small parameter, and $L$ is the total arc length of the body, and $\mean{\cdot}$ indicates a surface mean. The second term on the right acts to smooth bodies with high-frequency oscillations by introducing a {\em curvature}-dependent component in the interface velocity. Since a closed body has a mean curvature of $\mean{\kappa} = 2\pi / L$, the term in parentheses has mean zero. Thus, this term is area-preserving, so that the only material loss is due to the shear stress itself. In addition, we have scaled this term on the $\mean{\atau}$, the average absolute stress, so that it is always proportional to the true source of erosion $\atau$, even as bodies vanish (reword).

We now introduce the tangent angle, $\theta$, defined through the relation
\begin{equation}
\left( \pderiv{x}{s}, \pderiv{y}{s} \right) = \left(\cos \theta, \sin \theta \right)
\end{equation}
where $s$ is arc length. We also introduce the normalized arc length $\alpha = 2 \pi s / L$. In terms of these new variables, the curvature can be expressed as
\begin{equation}
\kappa = \frac{2 \pi}{L} \pderiv{\theta}{\alpha}
\end{equation}
Substitution into \eqref{eqn:Vn} gives
\begin{equation}
\label{eqn:Vn2}
\Vn = \atau +  \eps \mean{\atau}   \left(\pderiv{\theta}{\alpha} - 1 \right)
%\Vn = \atau + \frac{2 \pi \eps}{L \log \left(2\pi/L \right)}  \left(\pderiv{\theta}{\alpha} - 1 \right)
\end{equation}

The main idea is to introduce an artificial tangential velocity that keeps collocation points equally spaced with respect to arc length; i.e.~the grid of $\alpha$ should always remain equally spaced. The tangential velocity that does the trick is determined by \cite{hou-low-she1994}
\begin{equation}
\pderiv{\Vs}{\alpha} = \thalpha \Vn - \mean{\thalpha \Vn}
\end{equation}
The interface evolution is then given by
\begin{equation}
\dot{\xx}(t) = \Vn \nn + \Vs \ss.
\end{equation}
This evolution equation can be recast in the variables $\theta$ and $L$ as
\begin{align}
\label{eqn:Levo1}
& \tderiv{L}{t} = - 2 \pi \mean{\thalpha \Vn} \\
\label{eqn:thetaevo0}
& \pderiv{\theta}{t} = \frac{2 \pi}{L} \left( \pderiv{\Vn}{\alpha} + \thalpha \Vs \right)
\end{align}
Inserting \eqref{eqn:Vn2} for the normal velocity gives
\begin{equation}
\label{eqn:thetaevo1}
\pderiv{\theta}{t} = \frac{2\pi}{L} \left(
\eps \mean{\atau} \ppd{\theta}{\alpha} + \pderiv{\atau}{\alpha} + \thalpha \Vs \right)
\end{equation}
Note that the highest-order, i.e.~stiff, term in \eqref{eqn:thetaevo1} is {\em quasi-linear}. This is the key to the {\thL} method, and allows us to use very stable, implicit methods for this stiff term.

@@

%% TIME-STEPPING %%
\subsection{Time-stepping with the {\thL} method}

With the shear stress computed, we are now ready to evolve the interface of each body. As described in Section \ref{sec:thetaL}, we use the {\thL} framework, with $\theta$ being the tangent angle and $L$ the total arc length of a given boundary.

First, we introduce the variables
\begin{align}
\label{eq:M}
& \Mterm = - 2 \pi \mean{\thalpha \Vn} \\
\label{eq:N}
& \NLterm = \frac{2 \pi}{L} \left( \pderiv{\atau}{\alpha} + \thalpha \Vs \right) \\
\label{eq:elfun}
& \elfun = \frac{2 \pi  \mean{\atau}}{L}
\end{align} 
so that evolution equations \eqref{eqn:Levo1} and \eqref{eqn:thetaevo1} can be compactly expressed as
\begin{align}
\label{eqn:Levo2}
& \tderiv{L}{t} = \Mterm \\
\label{eqn:thetaevo2}
& \pderiv{\theta}{t} = \eps \elfun \ppd{\theta}{\alpha} + \NLterm
\end{align}
We will evolve $\theta$ in spectral space, and so introduce the transforms
\begin{align}
& \theta(\alpha,t) = \alpha + \FourierSum \thhat_k e^{ i k \alpha} \\
& \NLterm(\alpha,t)  = \FourierSum \hat{\NLterm}_k e^{ i k \alpha}
\end{align}
Recall that $\alpha = 2 \pi s / L$ is a normalized arc length. In spectral space, \eqref{eqn:thetaevo2} produces the system of ODEs
\begin{equation}
\tderiv{\thhat_k}{t} +  \eps k^2  \elfun \thhat_k = \hat{\NLterm}_k
\end{equation}
Each ODE is first-order and linear (though variable-coefficient due to $\elfun$) and thus can be solved by the integrating factor
\begin{equation}
\label{eq:mu}
\mu_k = \exp \left( \eps k^2 \int \elfun \, dt \right)
\end{equation}
We are therefore led to the coupled system of ODEs
\begin{align}
\label{eqn:Levo2}
& \tderiv{L}{t} = \Mterm \\
\label{eqn:thetaevo2}
& \tderiv{}{t}\left( \mu_k \thhat_k \right) = \mu_k \hat{\NLterm}_k
\end{align}
We solve system \eqref{eqn:Levo2}--\eqref{eqn:thetaevo2} with an explicit, second-order Runge-Kutta method (RK2). In particular, we choose the {\em midpoint} RK2 method so as to maximize the smoothing effect of the curvature term.

@@

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}[p]%[htbp]
\begin{center}
\includegraphics[width = 0.8 \textwidth]{./figs/arangle.pdf}
\caption{Aspect ratio and opening angle}
\label{}
\end{center}
\end{figure}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
% Figure from datafile 01circ1024c5
% Uses 01circ1024.in with epsfac = 5, sigfac = 5, dt = 2e-4, fixarea = 1, fixpdrop = 0

% Table
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{table}[p]%[htbp]
\begin{center}
\caption{Final aspect ratio and opening angle
} 
\vspace{0.3 pc}
%\label{aaa}
\begin{tabular}{c c c}
\hline
\hspace{0.5pc} smoothing parameters: $\eps$ and $\sigma$
\hspace{0.5pc} & final aspect ratio 
\hspace{0.5pc} & final opening angle \\
\hline
20/1024		& 2.55	& $110^\circ \pm 6^\circ$	\\
10/1024		& 2.62	& $104^\circ \pm 4^\circ$	\\
5/1024		& 2.65	& $102^\circ \pm 2^\circ$	\\
\hline
\end{tabular}
\end{center}
\end{table}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
% npts = 1024
